Testing Traversability of Protein Tunnels with Flexible Ligands Using Motion Planning

### 2017-06-30 paper.tex
jdu spat.

The examples of the color mapping are shown in Fig.~\ref{fig:properties}.
The accessibility (Fig.~\ref{fig:properties}a) shows that more than the half of the tunnel is not accessible (red part of the tunnel).
The throughput shows (Fig.~\ref{fig:properties}b) that the only difficult is the part around the bottleneck (red color in Fig.~\ref{fig:properties}b), and the second half of the tunnel is also traversible.

@@

Besides the color mapping, it is also necessary to show the computed trajectories.
%This is useful, for example, to find out in which part of the tunnel the trajectories detour.
Simple visualization of all trajectories could however be too slow for an interactive work.
Therefore, the trajectories are first clustered and then only the clusters are visualized.
Due to different lengths of the trajectories, they are first converted to a normalized form.
Let $\cstart$ represent the average starting position of all trajectories and let $d_{max}$ be the 3D Euclidean distance
of the most distant configuration from $\cstart$ amongst all trajectories.
A set of $M$ spheres centered at $\cstart$ are created with radii $r'_i=i {d_{max} \over M}$, where $i=0,\ldots, M-1$.
The trajectory $P=(q_1,\ldots,q_n)$ of length $n$ is represented by the normalized vector $v=(x_1,\ldots,x_M)$ of length $M$,
where $x_i$ is the 3D position of the nearest configuration $q \in P$ to the surface of the $i$-th sphere with the radius $r'_i$.
The distance between two normalized trajectories $v_i$ and $v_j$ is defined as
$d(v_i, v_j) = \frac{1}{M} \sum_{1 \leq k \leq N} |x_k^i x_k^j|$.
This distance is used in the UPGMA clustering technique~\cite{sokal1958statistical}.
The trajectories can be visualized using a representative of each cluster.
The number of trajectories in each cluster is represented by the width of the polyline (Fig.~\ref{fig:trajectories}).
%In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).

@@

The average runtimes, success rates and sizes of the built trees for the first tunnel are shown in Tab.~\ref{tab::rrt}.
The highest average success rate wasv achieved for the most scaled-down ligands ($\smin=0.3$) and it decreases with the increasing $\smin$. 
The difference between the minimal and maximal success rates indicates how the initial configurations of the ligand influence the ability to pass the tunnels.
For example, the most difficult initial configuration of $\LA$ and $\smin=0.3$ lead to success rate 78~\% (the easiest 
        initial configuration lead to success rate $100$~\%), but for the scale $\smin=0.4$, the worst success rate is
only $42$~\% and the best 96~\%.
The traversability of larger ligands ($\smin=0.4$ vs. $\smin=0.3$) is therefore more sensitive to the initial configuration.
This shows the importance of testing the traversability from multiple initial configurations.
The runtime is significantly higher for the $\LB$ ligand, which is caused by the larger number of conformations that need to be
examined in each expansion.

@@

Trajectories for $\LA$ in all detected tunnels were classified as successful if they reached the last sphere of the tunnel
to distance $\dt=2$~\AA and they were considered unsuccessful otherwise.
The throughput computed from the trajectories shows that the tunnels are difficult not only around bottlenecks, but
also in other places.
The comparison of the classic bottleneck (i.e., measured by the radius of spherical probe) and throughput is depicted in Fig.~\ref{fig::tunnel2}a.
The trajectories for $\smin=0.4$ are depicted in Fig.~\ref{fig::tunnel2}b and for $\smin=0.5$ in Fig.~\ref{fig::tunnel2}c.
The successful trajectories for $\smin=0.4$ reveals that the end of the tunnel No. 3 can be approached by two different pathways.
The detail of this pathway is depicted in Fig.~\ref{fig::detail}.
Despite the low bottleneck of this tunnel, the ligand may reach its end using the alternative path way. 
This shows the strength of the analysis based on the trajectories and not only on the geometric bottlenecks.

@@

\begin{figure}
\centering
\begin{tabular}{ccc}
\includegraphics[width=0.32\textwidth]{fig/tunne4ac}& 
\includegraphics[width=0.32\textwidth]{fig/tunne4bc} & 
\includegraphics[width=0.32\textwidth]{fig/tunne4cc} \\
(a) & (b) & (c)                        
\end{tabular}
\caption{\label{fig::detail}
    Detailed view to alternative pathway around the third tunnel in 1CQW.
    The successful trajectories are in green (a) the unsuccessful in red (b).
    (c) Shows visualization of protein atoms around the tunnel.
}
\end{figure}

@@

The second limitation is caused by the utilization of scaled-down ligands.
The protein atoms tend to fill the internal void space therefore the tunnels identified inside proteins without ligand tend
to be very narrow.
To enable at least some motion of the ligands inside such narrow tunnels, the scaling-down of the ligand atoms is necessary.
It is used also in other related tool, e.g.~\cite{cortes2005path}.
%By allowing more aggressive scaling-down, more easier a trajectory can be found.
%The tunnels therefore should be always evaluated considering various minimal scales.

Due to these reasons, the computed trajectories can be considered either as too optimistic
(e.g. because they are computed on a wide tunnel that can be in fact closed due to molecular dynamics),
or too pessimistic (e.g. the trajectory is not found because of narrow passage around the tunnel, which can be 
opened due to the molecular dynamics).
Despite these limitations, testing the tunnel traversability using motion planning technique can provide chemists more information
than the simple bottleneck radius, which is used nowadays.
For example, possible detours from the tunnel can be identified, which suggest that even a tunnel with a small
bottleneck can be traversed (Fig.~\ref{fig::detail}).

@@

Protein tunnels are transporting pathways for ligands towards active sites.
In this paper, we propose to analyze the traversability in the tunnels using motion planning, that can consider both shape
of the protein and ligand.
We proposed modifications to the RRT method to compute trajectories for a ligand represented by a library of common conformations.
The corresponding configuration space is sampled using the guided sampling, where a virtual sphere moves along the tunnel and attracts growth
of the tree towards it.
To enable trajectory computation in narrow tunnels, atomic radii of the ligand are scaled down.
Computed trajectories allows us to evaluate the traversability of the tunnels using
accessibility, throughput and also by visualizing typical pathways found by the planner.
These properties can help to better understand importance of the tunnels.

>>>>

The examples of the color mapping are shown in Fig.~\ref{fig:properties}.
The accessibility (Fig.~\ref{fig:properties}a) shows that more than the half of the tunnel is not accessible (red part of the tunnel).
The throughput shows (Fig.~\ref{fig:properties}b) that the only difficult is the part around the bottleneck (red color in Fig.~\ref{fig:properties}b), and the second half of the tunnel is also traversable.

@@

Besides the color mapping, it is also necessary to show the computed trajectories.
%This is useful, for example, to find out in which part of the tunnel the trajectories detour.
Simple visualization of all trajectories could however be too slow for an interactive work.
Therefore, the trajectories are first clustered and then only the clusters are visualized.
Due to different lengths of the trajectories, they are first converted to a normalized form.
Let $\cstart$ represent the average starting position of all trajectories and let $d_{max}$ be the 3D Euclidean distance
of the most distant configuration from $\cstart$ among all trajectories.
A set of $M$ spheres centered at $\cstart$ are created with radii $r'_i=i {d_{max} \over M}$, where $i=0,\ldots, M-1$.
The trajectory $P=(q_1,\ldots,q_n)$ of length $n$ is represented by the normalized vector $v=(x_1,\ldots,x_M)$ of length $M$,
where $x_i$ is the 3D position of the nearest configuration $q \in P$ to the surface of the $i$-th sphere with the radius $r'_i$.
The distance between two normalized trajectories $v_i$ and $v_j$ is defined as
$d(v_i, v_j) = \frac{1}{M} \sum_{1 \leq k \leq N} |x_k^i x_k^j|$.
This distance is used in the UPGMA clustering technique~\cite{sokal1958statistical}.
The trajectories can be visualized using a representative of each cluster.
The number of trajectories in each cluster is represented by the width of the polyline (Fig.~\ref{fig:trajectories}).
%In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).

@@

The average runtimes, success rates and sizes of the built trees for the first tunnel are shown in Tab.~\ref{tab::rrt}.
The highest average success rate was achieved for the most scaled-down ligands ($\smin=0.3$) and it decreases with the increasing $\smin$. 
The difference between the minimal and maximal success rates indicates how the initial configurations of the ligand influence the ability to pass the tunnels.
For example, the most difficult initial configuration of $\LA$ and $\smin=0.3$ lead to success rate 78~\% (the easiest 
        initial configuration lead to success rate $100$~\%), but for the scale $\smin=0.4$, the worst success rate is
only $42$~\% and the best 96~\%.
The traversability of larger ligands ($\smin=0.4$ vs. $\smin=0.3$) is therefore more sensitive to the initial configuration.
This shows the importance of testing the traversability from multiple initial configurations.
The runtime is significantly higher for the $\LB$ ligand, which is caused by the larger number of conformations that need to be
examined in each expansion.
The average values of the runtimes, which are in the order of minutes, are negligible in the comparison of runtime of MD simulations.

@@

Trajectories for $\LA$ in all detected tunnels were classified as successful if they reached the last sphere of the tunnel
to distance $\dt=2$~\AA and they were considered unsuccessful otherwise.
The throughput computed from the trajectories shows that the tunnels are difficult not only around bottlenecks, but
also in other places.
The comparison of the classic bottleneck (i.e., measured by the radius of spherical probe) and throughput is depicted in Fig.~\ref{fig::tunnel2}a.
The trajectories for $\smin=0.4$ are depicted in Fig.~\ref{fig::tunnel2}b and for $\smin=0.5$ in Fig.~\ref{fig::tunnel2}c.
The successful trajectories for $\smin=0.4$ reveal that the end of the tunnel No. 3 can be approached by two different pathways (one in the tunnel and another one outside the tunnel).
The detail is depicted in Fig.~\ref{fig::detail}.
Despite the low bottleneck of this tunnel, the ligand may reach its end using the alternative pathway.

@@

\begin{figure}
\centering
{
\renewcommand{\tabcolsep}{0pt}
\begin{tabular}{ccc}
\includegraphics[width=0.325\textwidth]{fig/tunne4ac}& 
\includegraphics[width=0.325\textwidth]{fig/tunne4bc} & 
\includegraphics[width=0.325\textwidth]{fig/tunne4cc} \\
(a) & (b) & (c)                        
\end{tabular}
}
\caption{\label{fig::detail}
    Detailed view to alternative pathway around the third tunnel in 1CQW.
    The successful trajectories are in green (a) the unsuccessful in red (b).
    (c) Shows visualization of protein atoms around the tunnel.
}
\end{figure}

@@

The second limitation is caused by the utilization of scaled-down ligands.
The protein atoms fill the internal void space therefore the tunnels identified inside proteins without ligand tend
to be very narrow.
To enable at least some motion of the ligands inside such narrow tunnels, the scaling-down of the ligand atoms is necessary.
It is used also in other related tool, e.g.~\cite{cortes2005path}.
%By allowing more aggressive scaling-down, more easier a trajectory can be found.
%The tunnels therefore should be always evaluated considering various minimal scales.

Due to these reasons, the computed trajectories can be considered either as too optimistic
(e.g. because they are computed on a wide tunnel that can be in fact closed due to molecular dynamics),
or too pessimistic (e.g. the trajectory is not found because of narrow passage around the tunnel, which can be 
opened due to the molecular dynamics).
Despite these limitations, testing the tunnel traversability using motion planning technique can provide chemists more information
than the simple bottleneck radius, which is used nowadays.
For example, possible detours from the tunnel can be identified, which indicates that even a tunnel with a small
bottleneck can be traversed (Fig.~\ref{fig::detail}).

@@

The protein tunnels are transporting pathways for ligands to the  active sites.
In this paper, we propose to analyze the traversability in the tunnels using motion planning, that can consider both shape
of the protein and ligand.
We proposed modifications to the RRT method to compute trajectories for a ligand represented by a library of common conformations.
The corresponding configuration space is sampled using the guided sampling, where a virtual sphere moves along the tunnel and attracts growth
of the tree towards it.
To enable trajectory computation in narrow tunnels, atomic radii of the ligand are scaled down.
Computed trajectories allows us to evaluate the traversability of the tunnels using
accessibility, throughput and also by visualizing typical pathways found by the planner.
These properties can help to better understand importance of the tunnels.
